---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fuzzylink

<!-- badges: start -->
<!-- badges: end -->

The goal of the `fuzzylink` package is to allow users to merge datasets with fuzzy matches on identifying variables. Suppose, for example, you have the following two datasets: 

```{r, echo = FALSE}
library(tidyverse)
dfA <- tribble(~name, ~age,
               'Timothy B. Ryan', 28,
               'James J. Pointer', 40,
               'Jennifer C. Reilly', 32)

dfB <- tribble(~name, ~hobby,
               'Tim Ryan', 'Woodworking',
               'Jimmy Pointer', 'Guitar',
               'Jessica Renny', 'Camping')

```

```{r}
dfA
dfB
```

We would like a procedure that correctly identifies that the first two rows of `dfA` and `dfB` are likely to be matches, but not the third row. The `fuzzylink()` function can do so.

```{r, eval = FALSE}
df <- fuzzylink(dfA, dfB, by = 'name')
df
```

The procedure works by taking *pretrained text embeddings* from OpenAI's GPT-3 and constructing a measure of similarity for each pair of names. These similarity metrics are then used as predictors in a statistical model estimating the probability that two name pairs represent the same entity.

## Installation

You can install the development version of `fuzzylink` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("joeornstein/fuzzylink")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(fuzzylink)
## basic example code
```

Under the hood, the `fuzzylink()` function takes the following steps:

### Step 1: Embedding

```{r}
strings_A <- unique(dfA$name)
strings_B <- unique(dfB$name)
embeddings <- get_embeddings(unique(c(strings_A, strings_B)))

dim(embeddings)
rownames(embeddings)
```

### Step 2: Similarity Scores

```{r}
sim <- get_similarity_matrix(embeddings, strings_A, strings_B)
sim
```

### Step 3: Create a Training Set

```{r}
train <- get_training_set(sim)
train
```

### Step 4: Fit Model

```{r}
model <- glm(as.numeric(match == 'Yes') ~ sim, data = train)
train$match_prob <- predict(model, train, type = 'response')
train
```

### Step 5: Create Matched Dataset

```{r}
df <- sim |> 
  reshape2::melt() |> 
  set_names(c('A', 'B', 'sim'))

df$match_probability <- predict(model, df, type = 'response')

df |> 
  filter(match_probability > 0.65) |> 
  select(-sim) |> 
  right_join(dfA, by = c('A' = 'name')) |> 
  left_join(dfB, by = c('B' = 'name'))
```
